quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-10-15 paper-quollr.Rmd
Initial work adding paragraph for detourr implementation

In the resulting interactive visualization black points represent the high-dimensional data, green points represent the model centroids from each bin, and the lines between model points reflect the \twoD wireframe structure mapped back to high-dimensional space.

@@

>>>>

@@

As an alternative to `langevitour`, users can explore the fitted \pD model using the `detourr` [@casper2025]. The combined data object from `comb_data_model()` can be passed directly to the `detour()` function, where `tour_aes()` defines the projection variables and color mapping. The visualization is rendered using `show_scatter()`, which can display both data points and the model’s structural edges via the `edges` argument.

```{r, echo=TRUE, eval=knitr::is_html_output()}
library(detourr)

detour(
  df_exe,
  tour_aes(
    projection = starts_with("x"),
    colour = type
  )
) |>
  show_scatter(axes = TRUE, 
               edges = as.matrix(trimesh[, c("from_reindexed", "to_reindexed")]),
               palette = c("#66B2CC", "#FF7755"))
```

In the resulting interactive visualization, blue points represent the high-dimensional data, orange points represent the model centroids from each bin, and the lines between model points reflect the \twoD wireframe structure mapped to high-dimensional space.

### 2025-10-15 paper-quollr.tex
Initial work adding paragraph for detourr implementation

In the resulting interactive visualization black points represent the high-dimensional data, green points represent the model centroids from each bin, and the lines between model points reflect the $2\text{-}D$ wireframe structure mapped back to high-dimensional space.

@@

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-32-1} \caption{Exploring the correspondence between UMAP layout and `Scurve` structure in $7\text{-}D$. Two sets of plots are linked: UMAP layout (a1, b1) and projection of $7\text{-}D$ model and data (a2, b2). The purple points indicate the selected subsets, which differ between rows. In (a1), the lower bridge of the `Scurve` is highlighted, which corresponds in (a2) to points spanning across both arms of the high-dimensional structure. In (b1), a different region near the upper arm of the `Scurve` is selected, and in (b2) these points map onto one side of the curved manifold in $7\text{-}D$ projection. While the UMAP layout suggests distinct local clusters, the linked tour views reveal how these selections trace continuous structures in the $7\text{-}D$ space, highlighting distortions introduced by UMAP.}\label{fig:unnamed-chunk-32}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-36-1} \caption{Exploring residuals in relation to UMAP layouts using a $7\text{-}D$ `Scurve` model. Three views are linked: distribution of residuals (a1, b1), UMAP layout (a2, b2), and projection of the $7\text{-}D$ model with data (a3, b3). The purple points highlight selected subsets of the data, which differ across rows. In the top row (a1–a3), points with higher residuals (a1) are selected, corresponding to the sparse bridging region in the UMAP layout (a2) and the less dense end of the `Scurve` in the high-dimensional projection (a3). In the bottom row (b1–b3), points with lower residuals (b1) are highlighted, which map to one side of the dense region in the NLDR layout (b2) and to a thicker band of the `Scurve` in the projection (b3). This comparison illustrates how residuals can help diagnose distortions in UMAP, with high-residual points often concentrated in sparse or stretched regions of the structure.}\label{fig:unnamed-chunk-36}
\end{figure}

>>>>

@@

As an alternative to \texttt{langevitour}, users can explore the fitted $p\text{-}D$ model using the \texttt{detourr} \citep{casper2025}. The combined data object from \texttt{comb\_data\_model()} can be passed directly to the \texttt{detour()} function, where \texttt{tour\_aes()} defines the projection variables and color mapping. The visualization is rendered using \texttt{show\_scatter()}, which can display both data points and the model's structural edges via the \texttt{edges} argument.

\begin{verbatim}
library(detourr)

detour(
  df_exe,
  tour_aes(
    projection = starts_with("x"),
    colour = type
  )
) |>
  show_scatter(axes = TRUE, 
               edges = as.matrix(trimesh[, c("from_reindexed", "to_reindexed")]),
               palette = c("#66B2CC", "#FF7755"))
\end{verbatim}

In the resulting interactive visualization, blue points represent the high-dimensional data, orange points represent the model centroids from each bin, and the lines between model points reflect the $2\text{-}D$ wireframe structure mapped to high-dimensional space.

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-33-1} \caption{Exploring the correspondence between UMAP layout and `Scurve` structure in $7\text{-}D$. Two sets of plots are linked: UMAP layout (a1, b1) and projection of $7\text{-}D$ model and data (a2, b2). The purple points indicate the selected subsets, which differ between rows. In (a1), the lower bridge of the `Scurve` is highlighted, which corresponds in (a2) to points spanning across both arms of the high-dimensional structure. In (b1), a different region near the upper arm of the `Scurve` is selected, and in (b2) these points map onto one side of the curved manifold in $7\text{-}D$ projection. While the UMAP layout suggests distinct local clusters, the linked tour views reveal how these selections trace continuous structures in the $7\text{-}D$ space, highlighting distortions introduced by UMAP.}\label{fig:unnamed-chunk-33}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-37-1} \caption{Exploring residuals in relation to UMAP layouts using a $7\text{-}D$ `Scurve` model. Three views are linked: distribution of residuals (a1, b1), UMAP layout (a2, b2), and projection of the $7\text{-}D$ model with data (a3, b3). The purple points highlight selected subsets of the data, which differ across rows. In the top row (a1–a3), points with higher residuals (a1) are selected, corresponding to the sparse bridging region in the UMAP layout (a2) and the less dense end of the `Scurve` in the high-dimensional projection (a3). In the bottom row (b1–b3), points with lower residuals (b1) are highlighted, which map to one side of the dense region in the NLDR layout (b2) and to a thicker band of the `Scurve` in the projection (b3). This comparison illustrates how residuals can help diagnose distortions in UMAP, with high-residual points often concentrated in sparse or stretched regions of the structure.}\label{fig:unnamed-chunk-37}
\end{figure}
